# Inference performance {#Inference-performance}

```{css, echo=FALSE}
pre {
  max-height: 1000px;
  overflow-y: auto;
}

pre[class] {
  max-height: 500px;
}
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

One of the primary purposes of the `DAISIEmainland` package and specifically
why the data is formatted in the `DAISIE` format is to test the maximum 
likelihood inference models implemented in the `DAISIE` R package [@etienne_daisie_2022]. Therefore, in this section we explore how to conduct a
simple performance analysis of one of the `DAISIE` models. In this case we are
going to use the model with a single macroevolutionary regime on the island (
i.e. all island species are assumed to have the same rate of colonisation, 
speciation and extinction, as well as the same carrying capacity). This model
also assumes that the carrying capacity only operates between species within the
same island clade (termed clade-specific diversity-dependence), thus different
island colonists are supposed independent and not inhibiting the diversification
of each other.

As a small technical aside, this section uses `DAISIE` version 4.1.1. Re-running
this code on another system may produce different results, especially if a
different version of `DAISIE` is installed.

## Simulating data {#Simulating-data}

First, we simulate 100 replicates of island data. This will produce phylogenetic data for a 100 islands with the same parameter values. The reason multiple island
data sets are simulated is because the simualation algorithm is stochastic (Section \@ref(Simulation-algorithm)) and by iterating the simulation many times
it accounts for differences between data sets due to stochasticity. In an analysis more replicates (e.g. 1,000) can be run to account for the stochastic differences between replicates to more thoroughly sample the distribution of possible simulation outcomes.

Note: all the code below takes a substantial amount of time to run (on the order
of hours for 100 replicates). 

```{r, load simulated data, echo=FALSE}
daisie_mainland_data <- readRDS(
  file = system.file(
    "inst/book/data/daisie_mainland_data_100_rep.rds", 
    package = "DAISIEmainland")
)
```

```{r, run-simulation, eval=FALSE}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection")

replicates <- 100

daisie_mainland_data <- DAISIEmainland::sim_island_with_mainland(
  total_time = 1,
  m = 100,
  island_pars = c(0.5, 02.5, 50, 0.01, 0.5),
  mainland_ex = 1.0,
  mainland_sample_prob = 1,
  mainland_sample_type = "complete",
  replicates = replicates,
  verbose = FALSE
)
```

Now we have the simulated data, stored in `daisie_mainland_data` (see section \@ref(daisie-mainland-data)), which is a list object.
At the highest level of the list there are two lists: 
`daisie_mainland_data$ideal_multi_daisie_data` and
  `daisie_mainland_data$empirical_multi_daisie_data`. Each of these `multi_daisie_data` lists contains 100 elements, one for each simulation replicate. The `daisie_mainland_data$ideal_multi_daisie_data` and `daisie_mainland_data$empirical_multi_daisie_data` have the same structure.
The first element of each is the meta data, containing: `island_age` and the 
number of species `not_present` on the island that are on the mainland.

```{r Inspect data part 1}
daisie_mainland_data$ideal_multi_daisie_data[[1]][[1]]
daisie_mainland_data$empirical_multi_daisie_data[[1]][[1]]
```

Subsequent elements of the list are the island clades which are composed of:
`branching_times`, status of colonisation or `stac`, and number of
`missing_species`. The `branching_times` contains the age of the island, the
time of colonisation and any subsequent cladogenetic speciation times. The `stac`
is a numeric identifier of the endemicicity status and cladogenetic status (i.e. is the island colonist a singleton lineage or a clade). Lastly, the `missing_species` is the number of species known from an island clade but not
included in the `branching_times` vector as there is no phylogenetic information
on the timing of speciation for that species. In the `DAISIEmainland` simulation
we assume that we have phylogenetic information on all species on the island and thus `missing_species` is always set to zero.

```{r Inspect data part 2}
daisie_mainland_data$ideal_multi_daisie_data[[1]][[2]]
daisie_mainland_data$empirical_multi_daisie_data[[1]][[2]]
```

## Maximum likelihood inference {#Maximum-likelihood-inference}

To run a maximum likelihood DAISIE model, using the `DAISIE_ML_CS()` function, on each replicate we create objects to store the data in (`ideal_ml` and `empirical_ml`) and then loop over each replicate.

```{r, load ml data, echo=FALSE}
daisie_ml <- readRDS(
  file = system.file(
    "inst/book/data/daisie_ml_100_rep.rds", 
    package = "DAISIEmainland")
)

ideal_ml <- daisie_ml$ideal_ml
empirical_ml <- daisie_ml$empirical_ml
```

```{r Calculate parameter estimates, eval=FALSE}
ideal_ml <- vector("list", replicates)
empirical_ml <- vector("list", replicates)

for (i in seq_len(replicates)) {
  ideal_ml[[i]] <- DAISIE::DAISIE_ML_CS(
    datalist = daisie_mainland_data$ideal_multi_daisie_data[[i]],
    initparsopt = c(0.5, 02.5, 50, 0.01, 0.5),
    idparsopt = 1:5,
    parsfix = NULL,
    idparsfix = NULL,
    ddmodel = 11,
    methode = "odeint::runge_kutta_fehlberg78",
    optimmethod = "simplex",
    jitter = 1e-5)
  
  empirical_ml[[i]] <- DAISIE::DAISIE_ML_CS(
    datalist = daisie_mainland_data$empirical_multi_daisie_data[[i]],
    initparsopt = c(0.5, 02.5, 50, 0.01, 0.5),
    idparsopt = 1:5,
    parsfix = NULL,
    idparsfix = NULL,
    ddmodel = 11,
    methode = "odeint::runge_kutta_fehlberg78",
    optimmethod = "simplex",
    jitter = 1e-5)
}
```

The details of the maximum likelihood set up are not important, but a brief explainer: all model parameters are optimisied with the starting position in parameter space for optimisation, equal to the values used to simulate the data; the Runge-Kutta Fehlberg method to numerically solve the likelihood equations; and simplex is the optimisation algorithm to maximise the likelihood.

## Inference performance error metrics {#Inference-performance-error-metrics}

Now we have simulated 100 data sets, each with an *ideal* and *empirical* data
set, and fitted the DAISIE model to each, we need to quantify the error the DAISIE inference makes because it does not include mainland evolutionary dynamics. 

```{r, load error metrics, echo=FALSE}
error_metrics <- readRDS(
  file = system.file(
    "inst/book/data/error_metrics_100_rep.rds", 
    package = "DAISIEmainland")
)
```

The error metrics chosen to quantify this are:

1. The difference between the parameter estimates from *ideal* and *empirical* data, for cladogenesis (\$clado_ratio), extinction (\$ext_ratio), carrying capacity (\$k_ratio), colonisation (\$immig_ratio), and anagenesis (\$ana_ratio). This can be calculated as the *ideal* estimate minus *empirical* estimate,  or *ideal* estimate divided by the *empirical* estimate.

```{r Parameter differences, class.output="scroll-100"}
error_metrics$param_diffs
```

```{r Parameter ratios, class.output="scroll-100"}
error_metrics$param_ratios
```
  
2. Delta CTT ($\Delta$CTT) (difference in colonisations through time) between
the *ideal* and *empirical* data.

```{r Delta-CTT}
error_metrics$delta_ctt
```

3. Percentage of maximum island age colonisations (i.e. colonisations where
the most recent colonisation time extracted from the phylogenetic data is older
than the island) for the *ideal* and *empirical* data (only including colonisations that survive to the present). The ideal max age percentage is always zero as it is always known exactly when the species colonised the island, but is still calculated to check it is zero. The empirical max age percent can be any
percent [0, 100].

```{r Max-age-percentages}
error_metrics$max_age_percent
```

4. Percent of endemic species at the present. This includes counts of the 
number of endemic and non-endemic species in the *ideal* and *empirical* data, as 
well as the calculation of the percentage of endemic species in each data set.

```{r Endemic-percentages}
error_metrics$endemic_percent
```

Overall, all of these error metrics can be computed using `calc_error()` with the simulated data and the maximum likelihood estimates.

```{r Calculate error, eval=FALSE}
errors <- DAISIEmainland::calc_error(
  daisie_mainland_data = daisie_mainland_data,
  ideal_ml = ideal_ml,
  empirical_ml = empirical_ml
)
```

The code displayed in this chapter is a simplified version of the script used
to carry out the inference performance analysis for CITE BIORXIV which can be found at /scripts/run_analysis.R in the `DAISIEmainland` package. 
