# Installation {#installation}

You can install the released version of DAISIEmainland from github with:

```{r, eval=FALSE}
# install.packages("remotes") # install remotes package if not installed
remotes::install_github("joshwlambert/DAISIEmainland")
```

The `DAISIEmainland` package has several dependencies (the entire list can be
found in the DESCRIPTION file). Most of these are on CRAN and can be installed with
`install.packages()`. However, the `DAISIE` package [@etienne_daisie_2022] requires 
a more up-to-date version than available on CRAN and need to be installed from github:

```{r, eval=FALSE}
remotes::install_github("rsetienne/DAISIE")
```

Load the `DAISIEmainland` package with the `library()` function. This will also
load the package dependencies, specifically the `DAISIE` package which will be
used later on for parameter estimation by fitting maximum likelihood models.

```{r load package}
library(DAISIEmainland)
```

`DAISIEmainland` is currently not hosted on CRAN and thus cannot be installed
using `install.packages()`.

The package was developed using R version >= 4. Therefore, it is recommended to 
use an R version at least as recent as R 4.0.0. The package is tested using
continuous integration on the release version of R and the development version
of R. Therefore, any issues with running `DAISIEmainland` functions on older
versions of R will not be patched.
