# Simulation algorithm {#Simulation-algorithm}

The Doob-Gillespie algorithm is a stochastic exact solution that is used to
simulate continuous-time processes, with several applications 
in biological modelling. The Doob-Gillespie algorithm
can be used in evolutionary biology, for example to efficiently simulate a
birth-death process. The island-mainland simulation in the DAISIEmainland
package uses a two-part Doob-Gillespie simulation, one for the mainland
(`DAISIEmainland::sim_mainland`) and one for the island
(`DAISIEmainland::sim_island`).

## Mainland simulation {#Mainland-simulation}

The simulation of the island phylogenetic data takes two simulations. Firstly
the mainland, which is simulated under a Moran process, whereby every species extinction is immediately followed by a random species giving rise to two new species (speciation). Then the mainland phylogenetic data is fed into the 
island simulation (Section \@ref(Island-simulation)).

The mainland simulation uses a Doob-Gillespie algorithm to simulate the 
speciation and extinction of species under a Moran process to maintain a
constant number of species. Here the example shows mainland data being 
simulated with for a time duration of one (million years), five initial mainland species, and a mainland extinction rate of one (per species per million years). 

```{r}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
mainland <- DAISIEmainland:::sim_mainland(
  total_time = 1,
  m = 5,
  mainland_ex = 1
)
```

The output is a list of five mainland clades called `multi_mainland_clade` (
see Section \@ref(multi-mainland-clade)).

```{r}
mainland
```

Specifically focusing on one of these mainland clades (see Section \@ref(mainland-clade)):

```{r}
mainland[[1]]
```

We can see that the information on this mainland clade include the ID of each
species (`spec_id`). The ID of the mainland ancestor from which that species
derived (`main_anc_id`). The type of species (`spec_type`) these can be `I`, 
`C`, or `E` for single lineage clade, cladogenetic species in a clade, or 
extinct species, respectively. All mainland species are initialised as 
`spec_id = I` but as the simulation progresses species become either 
part of a clade (`C`) or extinct (`E`). By keeping the extinct species we have
a full history of the mainland and not the so-called reconstructed history which
would only contain the living (extant) species. The `branch_code` provides the
evolutionary relationships of the species in the clade and allows the topology
of the clade to be constructed. The `branch_t` is the time when the species 
speciated from their common ancestor (in time after the start of the simulation
and not time before present). The `spec_origin_t` is the time the species 
originated and is often equal to the `branch_t`. The `spec_ex_t` is the time the species went extinct (again in time after the start of the simulation). Species
that do not go extinct are give a `spec_ex_t` equal to the total time of the 
simulation. 

Another example of what can happen on the mainland is the extinction of a 
singleton lineage before it underwent speciation. 

```{r}
mainland[[4]]
```

## Island simulation {#Island-simulation}

The island simulation runs after the mainland simulation and uses the mainland
data to determine which species can immigrate to the island through time. Doob-Gillespie algorithm is altered to accommodate the dynamic 
mainland pool. The time-steps are bounded to not jump over changes on the 
mainland to ensure the present state of the system (i.e. species on mainland)
is always up-to-date. The algorithm checks whether any changes have occurs on
the mainland since the last time step and if so the system is updated and the
returned to the time at which the mainland last changed. This is valid owing to
the Markov (memoryless) property of the Doob-Gillespie algorithm.

```{r}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
island_tbl <- DAISIEmainland:::sim_island(
  total_time = 1,
  island_pars = c(1, 1, 10, 1, 1),
  mainland_clade = mainland[[1]],
  mainland_sample_prob = 1,
  mainland_sample_type = "complete"
)
island_tbl
```

The island simulation outputs a `island_tbl` (see Section \@ref(island-tbl)). 
The `island_tbl` includes again the species ID (`spec_id`), mainland ancestor
ID (`main_anc_id`), species type (`spec_type`), branching code (`branch_code`), and branching time (`branch_t`). It also includes the time at which the species
colonised the island (`col_t`) and the reason a species is anagenetic (i.e.
endemic to the island without being in an island clade). The reasons for a 
species to be anagenetic are: the mainland population of the species goes 
extinct (`mainland_extinction`), the species undergoes anagenesis on the island
(`immig_parent`), or the species formed a clade but all other members of the 
clade went extinct before the present (`clade_extinct`). 

One major difference between the island and mainland data that are produced by
`sim_island_with_mainland()` and `sim_island()` is that the latter only has 
information from the reconstructed point of view. This means extinct species 
are removed from the data set and only species extant at the end of the 
simulation are included.

For both the island and mainland Doob-Gillespie algorithms time steps are
sampled from an exponential distribution with rate:
  
  $$X = \lambda e ^{-\lambda x}, \text{ where } \lambda = \sum_j r_j$$
  
where $r_j$ are the rates, for the mainland process this is just the rate of
mainland extinction ($\mu_M$), this is the only mainland parameter, whereas,
for the island algorithm $r_j$ are the rates of cladogenesis ($\lambda^c$),
island extinction ($\mu$), colonisation ($\gamma$), and anagenesis
($\lambda^a$). After the time step ($\Delta$ t) is sampled the event is sampled
from a dynamic discrete probability distribution, weighted by its rate
(propensity) relative to all other rates: 
  
  $$r_i / \sum_j r_j$$

The system is then updated for the algorithm repeats until the time step exceeds
the total time of the simulation. 

The function that encapsulates and runs both of these simulations is `sim_island_with_mainland()`. This function also includes the formatting of the data and the assignment of an endemicity status to each island colonist, which 
needed by the `DAISIE` inference model. The `DAISIEmainland` simulation outputs two data sets: (1) contains full information of all species colonisation times, and (2) an incomplete information data set which resembles what an empirist would have access to (see Section \@ref(daisie-mainland-data)). These two data sets allow for the quantification of error in
estimation when the empirists does not have access to all the data.

```{r}
set.seed(
  1,
  kind = "Mersenne-Twister",
  normal.kind = "Inversion",
  sample.kind = "Rejection"
)
daisie_mainland_data <- sim_island_with_mainland(
  total_time = 1,
  m = 5,
  island_pars = c(1, 1, 10, 1, 1), 
  mainland_ex = 1,
  mainland_sample_prob = 1,
  mainland_sample_type = "unsampled",
  replicates = 1 
)
daisie_mainland_data
```
