# Summary and error metrics visualisation {#summary-error-metrics-visualisation}

There is a range of plotting functions to visualise the summary metrics and 
error metrics (Section \@ref(inference-performance-error-metrics)). 

For the plotting of simulation summary and error metrics we need to load the 
results from multiple different mainland extinction scenarios.

```{r, read-analysis-results}
analysis_results <- DAISIEmainland::read_analysis_results(
  data_folder_path = system.file(
    "/inst/book/data/param_sets/", 
    package = "DAISIEmainland"
  )
)
```

## Simulation summary metrics

First we run the simulation. This is the same simulation as in the inference
performance section (Section \@ref(simulating-data)).

```{r, echo=FALSE}
<<load-simulated-data>>
```

```{r, eval=FALSE}
<<run-simulation>>
```

The simulation summary metrics we can calculate for the *ideal* and *empirical* data are: 

1. number of species on the island at the end of the simulation.
2. number of colonisation events to the island (that survived to the present) at the end of the simulation.

```{r ideal-sim-num-spec}
ideal_sim_num_spec <- DAISIEmainland::calc_num_spec(
  multi_daisie_data = daisie_mainland_data$ideal_multi_daisie_data
)

ideal_sim_num_spec
```

```{r ideal-sim-num-col}
ideal_sim_num_col <- DAISIEmainland::calc_num_col(
  multi_daisie_data = daisie_mainland_data$ideal_multi_daisie_data
)

ideal_sim_num_col
```

```{r empirical-sim-num-spec}
empirical_sim_num_spec <- DAISIEmainland::calc_num_spec(
  multi_daisie_data = daisie_mainland_data$empirical_multi_daisie_data
)

empirical_sim_num_spec
```

```{r empirical-sim-num-col}
empirical_sim_num_col <- DAISIEmainland::calc_num_col(
  multi_daisie_data = daisie_mainland_data$empirical_multi_daisie_data
)

empirical_sim_num_col
```

We can plot the simulation summary metrics to view the mean, variance, maximum and 
minimum number of colonisations and number of species for the *ideal* and *empirical* 
data.

```{r, plot-simulation-summary-metrics}
plot_sim_metrics(
  analysis_results = analysis_results, 
  output_file_path = NULL
)
```

## Inference error metrics

The error metrics were run for a range of mainland extinction rates. Then, the 
error metrics can be plotted across different rates to determine how the error
varies with faster or slower mainland evolutionary dynamics.

Thus, plotting the $\Delta$CTT across mainland extinction rates between zero and two
(per species per million years) is informative. As explained in Section
\@ref(inference-performance-error-metrics) this metric is a comparison between
the *ideal* and *empirical* data sets produced by the `DAISIEmainland` simulation.

```{r, plot-delta-ctt}
plot_ctt_boxplot(
  analysis_results = analysis_results, 
  output_file_path = NULL,
  parameter = "mainland_ex"
)
```

The percentage of endemics on the island at the end of the simulation and 
percentage of maximum age colonisation times can be plotted for different values
of mainland extinction. Each of the boxplots below shows 100 island replicates.
The left hand panel is for the data set with complete information (*ideal*) and
the right hand panel is for the data set with incomplete information (*empirical*).

```{r, plot-endemics}
plot_endemics(
  analysis_results = analysis_results, 
  output_file_path = NULL, 
  parameter = "mainland_ex"
) 
```

```{r, plot-max-age}
plot_max_age(
  analysis_results = analysis_results, 
  output_file_path = NULL, 
  parameter = "mainland_ex"
)
```



The plots shown in this chapter are not an exhaustive list available from the 
package. For example, error metrics can also be plotted across different values of mainland sampling
probability by setting the `parameter = "mainland_sample_prob"` in the plotting
functions. The script used to make the plots for the two manuscripts using this
package can be found [here](https://github.com/joshwlambert/DAISIEmainland/blob/main/scripts/joss_data_vis.R) 
and [here](https://raw.githubusercontent.com/joshwlambert/DAISIEmainland/main/scripts/evolution_data_vis.R).
