# Summary and error metrics visualisation {#Summary-error-metric-visualisation}

There is a range of plotting functions to visualise the summary metrics and 
error metrics (Section\@ref(Inference-performance-error-metrics)). 

## Simulation summary metrics

First we run the simulation. This is the same simulation as in the inference
performance section (Section \@ref(Simulating-data)).

```{r, load-simulated-data, echo=FALSE}
<<load-simulated-data>>
```

```{r, eval=FALSE}
<<run-simulation>>
```

The simulation summary metrics we can calculate for the *ideal* and *empirical* data are: 
1. number of species on the island at the end of the simulation
2. number of colonisation events to the island (that survived to the present) at the end of the simulation

```{r ideal-sim-num-spec}
ideal_sim_num_spec <- DAISIEmainland::calc_num_spec(
  multi_daisie_data = daisie_mainland_data$ideal_multi_daisie_data
)

ideal_sim_num_spec
```

```{r ideal-sim-num-col}
ideal_sim_num_col <- DAISIEmainland::calc_num_col(
  multi_daisie_data = daisie_mainland_data$ideal_multi_daisie_data
)

ideal_sim_num_col
```

```{r empirical-sim-num-spec}
empirical_sim_num_spec <- DAISIEmainland::calc_num_spec(
  multi_daisie_data = daisie_mainland_data$empirical_multi_daisie_data
)

empirical_sim_num_spec
```

```{r empirical-sim-num-col}
empirical_sim_num_col <- DAISIEmainland::calc_num_col(
  multi_daisie_data = daisie_mainland_data$empirical_multi_daisie_data
)

empirical_sim_num_col
```

## Inference error metrics

```{r}
<<Delta-CTT>>
```

```{r, read-analysis-results}
analysis_results <- DAISIEmainland::read_analysis_results(
  data_folder_path = file.path("results")
)
```

```{r, plot-delta-ctt}
plot_ctt_boxplot(
  analysis_results = analysis_results, 
  output_file_path = NULL,
  parameter = "mainland_ex"
)
```

```{r}
<<Max-age-percentages>>
```

```{r}
<<Endemic-percentages>>
```


The plots shown in this chapter are not an exhaustive list available from the 
package. The script used to make the plots for the two manuscripts using this
package CITE JOSS and CITE BIORXIV can be found at /scripts/joss_data_vis.R and /scripts/evolution_data_vis.R, respectively, in the `DAISIEmainland` package.
